---
title: "week_2A_SQL_and_R"
author: "Brandon Chanderban"
format: html
editor: visual
execute: 
  echo: true
---

# Introduction

The principal objective of this week 2A assignment is to build a small movie rating dataset using PostgreSQL (within pgAdmin 4, where the data will be housed) and then connect the resulting database tables to Rstudio for analysis.

## Data Collection Plan

-   Six movies, recently released and relatively popular, will be selected (potentially using a platform like IMBD, which hosts user-generated ratings).

-   Five participants will then be consulted to garner their ratings of each of the selected movies (on a 1-5 review scale).

The collection of this data may be facilitated via a Google sheets form, with the link being shared in a group chat so participants can easily submit their responses.

## Movie-Rating Database Design

In the confines of pgAdmin 4, three relational tables will be concocted.

-   a table of 'users' (user_id SERIAL PRIMARY KEY, name TEXT)

-   a table of 'movies' (movie_id SERIAL PRIMARY KEY, title TEXT)

-   and a table of the actual 'ratings' (user_id INT REFERENCES users(user_id), movie_id INT REFERENCES movies(movie_id), rating INT CHECK (rating BETWEEN 0 and 5), PRIMARY KEY (user_id, movie_id))

### Notes:

The 'ratings' table will serve as the mechanism through which the many-to-many relationship between the 'users' and 'movies' tables are modeled. The composite primary key will serve to ensure that each user can rate each movie only once. For movies a reviewer has not seen, missing ratings will be recorded as 0 (as permitted by the CHECK constraint) and later converted to NA in R.

The tables themselves will be generated via 'CREATE TABLE' statements in pgAdmin 4, and their data will be inserted manually via 'INSERT INTO' commands. Once created and populated, 'JOIN' queries will be used to combine data across the three data tables for export and analysis.

# R Analysis Plan

The 'RPostgres' and 'DBI' packages will be used to connect to the PostgreSQL database. The ratings data will then be loaded into Rstudio as a data frame.

Using the 'dplyr' package, the following exploratory analyses will be conducted:

-   Calculating the average rating per movie

-   Counting the number of ratings submitted by each reviewer

-   Counting the number of ratings received per movie

-   Handling missing data (replacing 0s with NA)

# Optional Endeavors

-   Visualizing rating distributions with the 'ggplot2' library

-   Normalizing ratings by user to adjust for individual rating biases

# Code Base (Body)

To begin, the necessary packages for connecting the PostgreSQL database (via pgAdmin 4) to RStudio were installed:

```{r install-packages}

install.packages('DBI')
install.packages('RPostgres')

```

Subsequently, the connection to the database was established using the dbConnect() function.

```{r connect-databse}

library(dplyr)
library(DBI)
library(RPostgres)

# Connect to PostgreSQL
con <- dbConnect(
  RPostgres::Postgres(),
  dbname = "movie_ratings",     # Name of the created database
  host = "localhost",           # Assuming PostgreSQL is running locally
  port = 5432,                  # Default PostgreSQL port
  user = "postgres",            
  password = Sys.getenv("DB_PASSWORD")     
)

```

The connection was tested by listing the available tables within the database.

```{r connection-test}

dbListTables(con)

```

The next step entailed importing the relevant tables into R as data frames using SQL queries via dbGetQuery()

```{r importing-data-tables}

# Load users table
users <- dbGetQuery(con, "SELECT * FROM users")

# Load movies table
movies <- dbGetQuery(con, "SELECT * FROM movies")

# Load ratings table
ratings <- dbGetQuery(con, "SELECT * FROM ratings")

```

Finally, the three tables were joined to provide full context for the ratings, incorporating both user and movie information.

```{r joining-tables}

ratings_full <- ratings %>%
  left_join(users, by = c("user_id" = "user_id")) %>%
  left_join(movies, by = c("movie_id" = "movie_id"))

head(ratings_full, 7)
```

Now that the data is properly housed within an R data frame (ratings_full), the next step is to address instances of missing data in the 'rating' column. In this dataset, a value of 0 represents cases where a reviewer had not seen a particular movie. To handle this appropriately, these values are recoded as NA. This treatment preserves the integrity of the dataset while enabling more flexible analytical approaches in the future (for example, if a recommender system were to be constructed using this dataset).

```{r convert-missing-values}

ratings_full <- ratings_full %>%
  mutate(rating = na_if(rating, 0))

head(ratings_full)

```

# Exploratory Data Analysis

In this section, the first analytical task undertaken pertains to the calculation of the average rating for each movie.

```{r average-rating-per-movie}

ratings_full %>%
  group_by(title) %>%
  summarise(
    avg_rating = mean(rating, na.rm = TRUE),
    num_ratings = sum(!is.na(rating))
  ) %>%
  arrange(desc(avg_rating))

```

After computing the mean ratings by movie title, the results were ordered in descending order based on the average rating. From this summary, it is observed that *Predator: Badlands* received the highest average rating among reviewers, while *Now You See Me: Now You Don't* received the lowest.

When examining the number of ratings per movie, all five reviewers reported having seen both *Avatar: Fire and Ash* and *People We Meet On Vacation. Fire and Ash* proved to be a major theatrical release, topping the U.S. and Canadian box offices during its opening weekend, as reported by this [BBC news article.](https://www.bbc.com/news/articles/cly5jrrj227o) In contrast, *People We Meet On Vacation* was released via Netflix rather than through a traditional theatrical distribution, which may have contributed to its similarly high viewership among participants.

```{r ratings-per-user}

ratings_full %>%
  group_by(name) %>%
  summarise(total_rated = sum(!is.na(rating))) %>%
  arrange(desc(total_rated))

```

The number of ratings submitted by each participant was also examined. From this summary, it is observed that Krystol provided the greatest number of movie ratings, while Sunil provided the fewest. However, given the small sample size of movies (six titles), these differences may not be fully indicative of each participant's broader movie-viewing habits. With a larger and more diverse set of items, these patterns could be assessed more reliably.

# Optional Endeavor: Visualization

In this optional extension, a bar chart was created to visualize average movie ratings while simultaneously displaying the number of ratings received per movie. In this visualization, bar height represents the average rating, and the number of submitted ratings is displayed as a label at the end of each bar.

```{r ratings_visualization}

library(ggplot2)

# Summarize the data
ratings_summary <- ratings_full %>%
  group_by(title) %>%
  summarise(
    avg_rating = mean(rating, na.rm = TRUE),
    num_ratings = sum(!is.na(rating))
  ) %>%
  arrange(desc(avg_rating))

# Plot with number of ratings as text labels
ggplot(ratings_summary, aes(x = reorder(title, avg_rating), y = avg_rating)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = num_ratings), hjust = -0.1, size = 3.5) +  
  coord_flip() +
  labs(
    title = "Average Movie Ratings",
    subtitle = "Labeled with Number of Ratings per Movie",
    x = "Movie Title",
    y = "Average Rating"
  ) +
  ylim(0, 5.5) +  # Extend y-axis to give space for text labels
  theme_minimal()

```

The above visualization provides an intuitive comparison across movies, allowing both rating magnitude and participation levels to be assessed simultaneously.

# Conclusion

This assignment served to demonstrate a small-scale, end-to-end workflow for collecting, storing, and analyzing user-item data using PostgreSQL and R. Key findings include meaningful variation in movie ratings, differences in user participation, and the presence of missing data due to unreviewed movie titles, an issue commonly encountered in real-world survey and recommendation settings.

From a methodological standpoint, the use of a normalized relational schema and the recoding of missing ratings as NA provide a solid foundation for further analysis. The results suggest that while average ratings offer useful insight, they should be interpreted alongside the number of ratings received to avoid overemphasizing movies with fewer reviews.

Several avenues exist to extend and strengthen the scope of this work. Future iterations could expand the number of users and movies to improve generalizability, incorporate additional metadata (such as genre, release year, or production company), or apply user-level rating normalization to account for individual rating tendencies. Additionally, this dataset could serve as a starting point for building a simple recommender system.

# Appendix

## A: SQL Code Used To Create And Populate Tables

```{sql}

--Create the users table
CREATE TABLE users (
    user_id SERIAL PRIMARY KEY,
    name TEXT NOT NULL
);

--Create the movies table
CREATE TABLE movies (
    movie_id SERIAL PRIMARY KEY,
    title TEXT NOT NULL
);

--Create the ratings table
CREATE TABLE ratings (
    user_id INT NOT NULL,
    movie_id INT NOT NULL,
    rating INT CHECK (rating BETWEEN 0 AND 5),

    PRIMARY KEY (user_id, movie_id),

    CONSTRAINT fk_user
        FOREIGN KEY (user_id)
        REFERENCES users (user_id)
        ON DELETE CASCADE,

    CONSTRAINT fk_movie
        FOREIGN KEY (movie_id)
        REFERENCES movies (movie_id)
        ON DELETE CASCADE
);

--Insert users
INSERT INTO users (name) VALUES
  ('Brianna'),
  ('Krystol'),
  ('Nalini'),
  ('Sunil'),
  ('Destiny');

--Insert movies
INSERT INTO movies (title) VALUES
  ('Zootopia 2'),
  ('Avatar: Fire and Ash'),
  ('People We Meet On Vacation'),
  ('Predator: Badlands'),
  ('Tron: Ares'),
  ('Now You See Me: Now You Don''t');

-- Insert Brianna's ratings (user_id = 1)
INSERT INTO ratings (user_id, movie_id, rating) VALUES
  (1, 1, 3),
  (1, 2, 5),
  (1, 3, 5),
  (1, 4, 4),
  (1, 5, 0),
  (1, 6, 3);

-- Insert Krystol's ratings (user_id = 2)
INSERT INTO ratings (user_id, movie_id, rating) VALUES
  (2, 1, 4),
  (2, 2, 4),
  (2, 3, 1),
  (2, 4, 4),
  (2, 5, 3),
  (2, 6, 1);

-- Inserting Nalini's ratiings (user_id = 3)
INSERT INTO ratings (user_id, movie_id, rating) VALUES
  (3, 1, 3),
  (3, 2, 5),
  (3, 3, 5),
  (3, 4, 5),
  (3, 5, 4),
  (3, 6, 0);

-- Insert Sunil's ratings (user_id = 4)
INSERT INTO ratings (user_id, movie_id, rating) VALUES
  (4, 1, 0),
  (4, 2, 2),
  (4, 3, 4),
  (4, 4, 0),
  (4, 5, 4),
  (4, 6, 5);

-- Insert Destiny's ratings (user_id = 5)
INSERT INTO ratings (user_id, movie_id, rating) VALUES
  (5, 1, 4),
  (5, 2, 3),
  (5, 3, 2),
  (5, 4, 4),
  (5, 5, 0),
  (5, 6, 2);

```

## B: Closing The Connection

```{r close-database-connection}

# Close the PostgreSQL database connection
dbDisconnect(con)
```

## LLM Used

-   OpenAI. (2026). *ChatGPT (Version 4o)* \[Large language model\]. [https://chat.openai.com](https://chat.openai.com/) . Accessed February 07, 2026.
