---
title: "week_9_nyt_api"
author: "Brandon Chanderban"
date:   "March 26, 2026"
format:
  html:
    toc: true
    code-fold: show
    theme: flatly
execute:
  echo:    true
  warning: false
---

# Introduction/Approach

The objective of this assignment is to become familiar with the process of interacting with a public API within R, specifically the New York Times Developer Network, and to demonstrate how data retrieved from an API may be parsed and transformed into a usable data frame for analytical endeavors. In this case, the New York Times Books API will be utilized, which provides access to current and historical best-seller lists. The analysis will be centered around examining author presence within the lists, with the prevailing question being: **Which authors appear most frequently on the New York Times best-seller list within a selected category?**

## Proposed Plan

The analytical approach will likely follow the steps outlined below.

Firstly, an account will be created on the New York Times Developer Network in order to obtain a valid API key. This key will then be stored securely as an environment variable to ensure that it is not hard-coded within the script.

Subsequently, a request will be made to the Books API endpoint, specifically the current best-seller list endpoint, by constructing the appropriate URL and passing the API key as a parameter.

Once the request has been made, the returned JSON response will be imported into R and parsed using the jsonlite package. Given that the Books API response contains nested elements, particularly within the "results" and "books" fields, the relevant components will then be extracted and flattened into the desired tabular structure.

Following this, the extracted data will be converted into a tidy data frame, retaining only the variables of interest such as title, author, rank, and weeks on the list. Any inconsistencies or missing values will have to be identified and addressed during this part of the transformation process.

Finally, the resulting data frame will be used to perform a simple aggregation in order to determine which authors appear most frequently within the selected best-seller list, thereby answering the afore-posed research question.

# Code Base/Body

After an account was created for the NYT Developer Network, a novel app was created (data_607_week_nine) within which the Books API was enabled. Upon saving the app, we are now able to access an API key that will facilitate our linkage to the NYT Books API. This API key was then stored as an environment variable, which was consistent with both the assignment directives and standard practice.

Proceeding, the necessary libraries will be loaded in.

```{r load-libraries}

library(tidyverse)
library(jsonlite)
library(gt)

nyt_api_key <- Sys.getenv("NYT_API_KEY")
```

## Select the API Endpoint and Request

For the scope of this assignment, the selected Books API will be the current Hardcover Fiction best-seller list. The Books API documentation describes the list detail service as a sort of request made to the list endpoint, using the list name and either a date or the term *current*.

```{r construct-the-api-request, echo=TRUE, results='hide'}

nyt_books_url <- paste0(
  "https://api.nytimes.com/svc/books/v3/lists/current/hardcover-fiction.json?api-key=", nyt_api_key
)
```

The request can now be made, after which the derived JSON response can be imported into RStudio.

```{r import-json-response}

nyt_books_raw <- fromJSON(nyt_books_url)

glimpse(nyt_books_raw)
head(nyt_books_raw)
```

## Parsing the JSON Response

The Books API returns a nested JSON structure, so the relevant fields must now be explored and extracted from the response.

```{r inspect-response-structure}

names(nyt_books_raw)
names(nyt_books_raw$results)
```

Now, the most relevant fields pertaining to the goal of this analysis will be extracted from the books list. These are: title, author, rank, publisher, weeks on list, the books' ISBN numbers, as well as the book descriptions.

```{r parse-boks-to-a-tidy-dataframe}

nyt_books_df <- nyt_books_raw$results$books %>%
  as_tibble() %>%
  select(
    rank,
    title,
    author,
    publisher,
    weeks_on_list,
    primary_isbn13,
    description
  ) %>%
  rename(isbn13 = primary_isbn13)

glimpse(nyt_books_df)
head(nyt_books_df)
```

At this stage, the Books API response has been converted into a tidy data frame, with one row corresponding to one book presently on the selected best-seller list.

In considering the prior data cleaning decisions that were executed, the initial returned JSON response was nested, so only the desired fields from the book component were extracted into the final data frame. Secondly, the ISBN field was renamed to "isbn13" for additional clarity. We will now move onto the analytical portion of this assignment.

## Analysis: Which Author/s Appear Most Frequently on the Selected Best-Seller List

As set out within the former sections, the guiding question for this assignment is as follows: **Which authors appear most frequently on the selected best-seller list?** This selected list, of course, being the current best-seller list of Hardcover Fiction novels. Since the endpoint returned the current books on a single list, our task entails the counting the number of appearances of each author within said list.

```{r count-frequency-of-authors}

author_frequency <- nyt_books_df %>%
  count(author, sort = TRUE)

author_frequency
```

A summary table of the above data frame may also be constructed for easier readability.

```{r construct-author-frequency-summary-table}

author_frequency %>%
  gt::gt() %>%
  gt::tab_header(
    title = "Author Frequency on the Current Hardcover Fiction List",
    subtitle = "Count of Appearances by Author"
  ) %>%
  gt::cols_label(
    author = "Author",
    n = "Number of Appearances"
  )
```

And finally, the results may also be visualized by way of a simple bar chart.

```{r create-author-frequency-bar-plot}

ggplot(author_frequency, aes(x = reorder(author, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Author Frequency on the Current Hardcover Fiction List",
    x = "Author",
    y = "Number of Appearances"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

## Additional Analysis: Exploring Matt Dinniman's Books

Since Matt Dinniman was the author appearing the most times within the imported best-seller list for Hardcover Fiction, a further step in the analytical process could (and will) entail examining his book titles.

```{r examine-top-author}

top_author <- author_frequency %>%
  slice(1) %>%
  pull(author)

top_author_books <- nyt_books_df %>%
  filter(author == top_author) %>%
  select(title, weeks_on_list, description) %>%
  arrange(desc(weeks_on_list))

top_author_books
```

As done prior, for the frequency of author appearance on the best-seller list, the above data frame pertaining to Matt Dinniman's books may also be presented in a cleaner-looking table display.

```{r top-author-table}

top_author_books %>%
  gt::gt() %>%
  gt::tab_header(
    title = paste("Books by", top_author, "on the Current List"),
    subtitle = "Titles, Weeks on List, Descriptions"
  ) %>%
  gt::cols_label(
    title = "Title",
    weeks_on_list = "Weeks on List",
    description = "Descriptions"
  )
```

## Brief Interpretation

From the frequency table and the corresponding plot, it is evident that most authors appear only once on the current Hardcover Fiction best-seller list. However, *Matt Dinniman* stands out with multiple appearances (four) indicating a notably stronger presence within the list at this point in time. Upon closer examination of his entries, it can be observed that all four books belong to his *Dungeon Crawler Carl* series. Interestingly, the number of weeks each book has spent on the best-seller list appears to correspond with the order of the books within the series, with the earlier installments remaining on the list for longer durations than the more recent entries. This suggests a pattern of sustained reader engagement with the series over time.

# Conclusion

In completing this assignment, the New York Times Books API was successfully accessed through an authenticated request using an environment-stored API key, and the returned JSON response was parsed into a tidy data frame suitable for analysis. From this, it became possible to examine author representation within the current Hardcover Fiction best-seller list and identify a meaningful pattern within the extracted data. Overall, the exercise demonstrated how API-derived JSON data may be securely retrieved, cleaned, and transformed within RStudio for downstream analysis.

## LLM Used

-   OpenAI. (2026). *ChatGPT (Version 4o)* \[Large language model\]. [https://chat.openai.com](https://chat.openai.com/) . Accessed March 28, 2026.
